---
title: "code"
author: "Sorawan Tiratrakoonwichaya"
date: "2024-11-04"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#open raw data and clean data into R

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
raw_data <- read_delim("../raw_data/PlateletHW.tsv", delim = "\t", escape_double = FALSE, trim_ws = TRUE)
View(raw_data)

clean_data <- read_delim("../clean_data/clean_data.tsv", delim = "\t", escape_double = FALSE, trim_ws = TRUE)
View(clean_data)

```

#check for outliers in data
```{r}
summary(raw_data$ADP)
hist(raw_data$ADP, main = "Histogram from PlateletHW" , xlab = "ADP", col = "lightblue")
```
It has outliers because it has data lower than 0.
Therefore, the data must be corrected.

#create clean data
```{r}
clean_data$ADP <- abs(raw_data$ADP)
par(mfcol=c(1,2))
boxplot(raw_data$ADP, main = "Before")
boxplot(clean_data$ADP, main = "After")
```
```{r}
summary(clean_data$ADP)
hist(clean_data$ADP, main = "Histogram from clean_data" , xlab = "ADP", col = "salmon")
```



#Install car package
```{r}
install.packages("car")
```

# Checking The Relationship Between ADP Levels and Resistance, rs4244285, rs4986893, and rs662 Genotype
```{r}
library(car)
scatterplot(Resistance ~ ADP, data = clean_data, reg.line = lm, smooth = FALSE, main = "Relationship Between ADP Levels and Resistance")
boxplot(clean_data$ADP ~ clean_data$rs4244285, main = "Boxplot of ADP Levels by rs4244285 Genotype", xlab = "rs4244285", ylab = "ADP", col = "turquoise")
boxplot(clean_data$ADP ~ clean_data$rs4986893, main = "Boxplot of ADP Levels by rs4986893 Genotype", xlab = "rs4986893", ylab = "ADP", col = "slateblue")
boxplot(clean_data$ADP ~ clean_data$rs662, main = "Boxplot of ADP Levels by rs662 Genotype", xlab = "rs662", ylab = "ADP", col = "khaki")
```

# Normalize ADP by taking log 
```{r}
clean_data$ADP_log <- log(clean_data$ADP)
View(clean_data)
```


#Linear regression test
##ADP and rs4244285
```{r}
library(ggplot2)
linear_1 <- lm(ADP ~ rs4244285, data = clean_data)
summary(linear_1)# significant (p-value < 0.05)

ggplot(clean_data, aes(x = rs4244285, y = ADP)) +
  geom_point() +
  geom_smooth(method = "lm", color = "turquoise") +
  labs(title = "Linear regression test between ADP Level and rs4244285",
       x = "rs4244285 Genotype (0, 1, 2)",
       y = "ADP") +
  theme_minimal()

plot(linear_1$fitted.values, linear_1$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "turquoise")

qqnorm(linear_1$residuals, main = "QQ Plot: Relationship of ADP and rs4244285 (Unnormalized ADP)")
qqline(linear_1$residuals, col = "turquoise")

hist(linear_1$residuals, main = "Histogram of Residuals linear_1", xlab = "Residuals linear_1", col = "turquoise")

```

###If use the normalize ADP
```{r}
linear_log_1 <- lm(ADP_log ~ rs4244285, data = clean_data)
summary(linear_log_1)# significant (p-value < 0.05)

ggplot(clean_data, aes(x = rs4244285, y = ADP_log)) +
  geom_point() +
  geom_smooth(method = "lm", color = "turquoise") +
  labs(title = "Linear regression test between log(ADP Level) and rs4244285",
       x = "rs4244285 Genotype (0, 1, 2)",
       y = "log(ADP)") +
  theme_minimal()

plot(linear_log_1$fitted.values, linear_log_1$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "turquoise")

qqnorm(linear_log_1$residuals, main = "QQ Plot: Relationship of ADP and rs4244285 (Normalized ADP)")
qqline(linear_log_1$residuals, col = "turquoise")

hist(linear_log_1$residuals, main = "Histogram of Residuals linear_log_1", xlab = "Residuals linear_log_1", col = "turquoise")
```

##ADP and rs4986893
```{r}
linear_2 <- lm(ADP ~ rs4986893, data = clean_data)
summary(linear_2)# significant (p-value < 0.05)

ggplot(clean_data, aes(x = rs4986893, y = ADP)) +
  geom_point() +
  geom_smooth(method = "lm", color = "slateblue") +
  labs(title = "Linear regression test between ADP Level and rs4986893",
       x = "rs4986893 Genotype (0, 1, 2)",
       y = "ADP") +
  theme_minimal()

plot(linear_2$fitted.values, linear_2$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "slateblue")

qqnorm(linear_2$residuals, main = "QQ Plot: Relationship of ADP and rs4986893 (Unnormalized ADP)")
qqline(linear_2$residuals, col = "slateblue")

hist(linear_2$residuals, main = "Histogram of Residuals linear_2", xlab = "Residuals linear_2", col = "slateblue")

```

###If use the normalize ADP
```{r}
linear_log_2 <- lm(ADP_log ~ rs4986893, data = clean_data)
summary(linear_log_2)# significant (p-value < 0.05)

ggplot(clean_data, aes(x = rs4986893, y = ADP_log)) +
  geom_point() +
  geom_smooth(method = "lm", color = "slateblue") +
  labs(title = "Linear regression test between log(ADP Level) and rs4986893",
       x = "rs4986893 Genotype (0, 1, 2)",
       y = "log(ADP)") +
  theme_minimal()

plot(linear_log_2$fitted.values, linear_log_2$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "slateblue")

qqnorm(linear_log_2$residuals, main = "QQ Plot: Relationship of ADP and rs4986893 (Normalized ADP)")
qqline(linear_log_2$residuals, col = "slateblue")

hist(linear_log_2$residuals, main = "Histogram of Residuals linear_log_2", xlab = "Residuals linear_log_2", col = "slateblue")

```

##ADP and rs662
```{r}
linear_3 <- lm(ADP ~ rs662, data = clean_data)
summary(linear_3)# not significant (p-value > 0.05)

ggplot(clean_data, aes(x = rs662, y = ADP)) +
  geom_point() +
  geom_smooth(method = "lm", color = "khaki") +
  labs(title = "Linear regression test between ADP Level and rs662",
       x = "rs662 Genotype (0, 1, 2)",
       y = "ADP") +
  theme_minimal()

plot(linear_3$fitted.values, linear_3$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "khaki")

qqnorm(linear_3$residuals, main = "QQ Plot: Relationship of ADP and rs662 (Unnormalized ADP)")
qqline(linear_3$residuals, col = "khaki")

hist(linear_3$residuals, main = "Histogram of Residuals linear_3", xlab = "Residuals linear_3", col = "khaki")

```

###If use the normalize ADP
```{r}
linear_log_3 <- lm(ADP_log ~ rs662, data = clean_data)
summary(linear_log_3)# not significant (p-value > 0.05)

ggplot(clean_data, aes(x = rs662, y = ADP_log)) +
  geom_point() +
  geom_smooth(method = "lm", color = "khaki") +
  labs(title = "Linear regression test between log(ADP Level) and rs662",
       x = "rs662 Genotype (0, 1, 2)",
       y = "log(ADP)") +
  theme_minimal()

plot(linear_log_3$fitted.values, linear_log_3$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "khaki")

qqnorm(linear_log_3$residuals, main = "QQ Plot: Relationship of ADP and rs662 (Normalized ADP)")
qqline(linear_log_3$residuals, col = "khaki")

hist(linear_log_3$residuals, main = "Histogram of Residuals linear_log_3", xlab = "Residuals linear_log_3", col = "khaki")

```

# Check SEX and AGE association

```{r}
linear_sex <-lm(ADP_log ~ SEX, data= clean_data)
linear_age <- lm(ADP_log ~ AGE, data= clean_data)


summary(linear_sex)# not significant (p-value > 0.05)
summary(linear_age) # not significant (p-value > 0.05)

ggplot(clean_data, aes(x =SEX , y = ADP_log)) +
  geom_point() +
  geom_smooth(method = "lm", color = "orchid") + 
  labs(title = "Association between log(ADP) and SEX",
       x = "SEX 0 =male. 1 =SEX ",
       y = "log(ADP)") +
  theme_minimal()


ggplot(clean_data, aes(x =AGE , y = ADP_log)) +
  geom_point() +
  geom_smooth(method = "lm", color = "gold") + 
  labs(title = "Association between log(ADP) and AGE",
       x = "AGE",
       y = "log(ADP)") +
  theme_minimal()
```



# Create complex linear regression

```{r}
snp_list <- c("rs4244285", "rs4986893", "rs662")

results_list <- list()

for (snp in snp_list) {
  model_sum <- lm(as.formula(paste("ADP_log ~ AGE + SEX +", snp)), data = clean_data)
  results_list[[snp]] <- summary(model_sum)
}
print(results_list[["rs4244285"]]) # significant (p-value < 0.05/3)
print(results_list[["rs4986893"]])# significant (p-value < 0.05/3)
print(results_list[["rs662"]]) # not siginifcant (p-value > 0.05/3)
```



